Manuscript

### 2018-04-27 manuscript/manuscript.Rmd
added text

Why is human mobility important? How people move about in their environment affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. Therefore it is unsurprising that policymakers and social scientists in numerous fields are interested in human mobility measures. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@ipip]. Economists study geographic labour mobility extensively [e.g. @tatsiramos] and the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission]. Researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive epsiode in bipolar depression patients [@palmius_detecting_2017]. Indeed, within psychology when studying behavioural differences within individuals it has been often been neglected that individuals vary not only over time but also over space. To fully understand behaviour we must understand how behaviour can vary accross environments.

Despite their importance, a large part of studies on human mobility are conducted with pen-and-paper travel diaries. Pen-and-paper diares have well-known methodological flaws. For instance, they are expensive to collect and present a high burden to respondents. This limits the duration of data collection and the span of the log accordingly. Another issue is that they rely on self-reported measures. It has been reported that short trips are frequently under-reported  [@wolf_impact_2003] and the self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017]. These obstacles can be overcome by using objective data on human mobility. Such data  can now be obtained using the Global Positioning System (GPS). GPS uses the distance between a device and several satellites to determine location. GPS measurements can be used to infer a vast range of socioeconomic and behavioural measures, including where the individual lives, how much time he or she spends at home and where (and how) they travel.

However, not all GPS measurements are made equal. It is useful to classify GPS logs into different categories depending on their origin because different types of logs pose different methodological problems. We argue that social science researchers should explore using automatically collected smartphone GPS logs.  In most studies participants receive a specialised GPS devices to track their movement. We call resulting logs *specialised logs*. @barnett_inferring_2016 point out several methodological issues with these studies. Collecting specialised logs is costly and places a high burden on participants. Besides, introducing a new device to the participant's life may bias their behaviour. Due to these drawbacks specialised logs usually span a short amount of time. @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). The problem with this is that researchers have to develop a custom-made app, distribute it among research participants and enforce participation. Participants tend not to like these tracking apps because they drain the battery of the device. Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) . For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017]. Often, secondary logs span several years. By law, secondary logs are accessible to users for free [@commission_protecting_2017]. Yet, secondary logs also present methodological challenges. They were created for non-academic purposes under engineering constraints ( these constraints will be detailed subsequently). These constraints mean that sensors do not track users continuously, meaning that the resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with missing data and measurement noise.

@@

Another methodological problem is related to the noise in the measurements that are collected. The accuracy of smartphone location measurements is substantially lower than that of professional GPS trackers. Android phones collect location information through a variety of methods. Other than GPS measurements, Androids use less-accurate heuristics such as WiFi access points and cellphone triangulation. Different methods are used because of computational and battery constraints. GPS is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time.

@@

This paper wishes to achieve four objectives: First, we have argued that understanding human mobility is important. Secondly, we argued that social scientists should leverage data logs from smartphones to study human mobility, instead of relying on out-dated pen-and-paper questionnaires. Third, explained the practical difficulties in using smartphone location logs. In the following sections we would like to introduce Personal Map Matched Imputation (PPMI), a method for surmounting these difficulties. We will compare PPMI to existing methods in the literature.

>>>>

Why is human mobility important? How people move about in their environment affects a wide range of outcomes, such as health, income and social capital [@goodchild_toward_2010]. Therefore it is unsurprising that social scientists in numerous fields and even policymakers are interested in human mobility measures. The most widely administered personality questionnaires ask individuals to what extent they agree with statements such as "I love large parties", "I prefer going to the movies to watching videos at home" and "I love to travel to places that I have never been before" [@ipip]. In economic research, the postal code of an individual's home address is often used as a proxy for socioeconomic status [e.g. @villanueva]. Moreover, economists study geographic labour mobility extensively [e.g. @tatsiramos]. In fact, the European Commission considers labour mobility important enough to warrant its own directorate [@europeanComission]. The extensive use of measures like these across different domains strongly suggests that mobility metrics are linked to real world outcomes. Perhaps it is unsurprising that behavioural researchers have found that mobility measures can be used to predict academic performance [@wang_smartgpa:_2015], the incidence of obesity [@zenk_how_2009] or even the onset of a depressive epsiode in bipolar depression patients [@palmius_detecting_2017]. Indeed, there is an argument to be made that perhaps psychologists have not been studying mobility enough. When studying behavioural differences within individuals behavioural scientists have often neglected the fact that individuals vary not only over time but also over space. To fully understand behaviour we must understand how behaviour can vary accross environments.

Despite the importance of mobility measures, the majority of these measures are obtained through the use of questionnaires (such as the IPIP) or specifically through pen-and-paper travel diaries. Questionnaires and travel diaries are have well-known methodological flaws: they are burdensome to collect and rely on accurate self-reporting. The burdensomeness stems from the fact that participans must be explicitly asked to report on their movement patterns at frequent intervals due to forgetfulness. This makes the data expensive to collect and limits the extent of data collection in practice. In addition, the frequent reporting duties of the participant may bias the participants behaviour. Participant forgetfulness also limits the accuracy of self reported measures. There is clear evidence that participants are systematically biased when self-reporting mobility measures. For instance, participants underreport the frequency of short trips [@wolf_impact_2003] and underestimate the duration of regular commutes [@delclos-alio_keeping_2017]. These obstacles can be overcome by using objective data on human mobility.

Social scientists now have the unprecedented opportunity to easily obtain objective data on human mobility from smartphones. Before smartphones the only way to collect objective data on human mobility involved giving participants an expensive professional-grade location sensor and convincing them to take it with themselves at all times. @barnett_inferring_2016 points out that introducing a new device to the participant's life may bias their behaviour. Moreover, collecting data in such a way is costly, places a high burden on participants and therefore the logs do not span a long time [@barnett_inferring_2016. Today millions of individuals carry smartphones with themselves every day and do not need to be encouraged to do so by researchers. Smartphones are equipped with a range of sensors that can be used to track the location of the device at all times. These smartphones can collect and store hundreds of location measurements a day. For instance,  Google Location History contains movement information on millions of users, often spanning years [@location_history_timeline_2017]. Moreover, recent changes in EU regulations with regard to consumer data-portability rights ensure that a willing participant should be able to easily share this information with researchers at no cost to the either the participant or the researchers [@commission_protecting_2017]. Taken together, this means that researchers now have at their disposal the ability to easily access the objective human mobility data of millians of individuals spanning several years at little cost and without a significant burden of the participants.

This paper wishes to achieve four objectives: First, we have argued that understanding human mobility is important. Secondly, we argued that social scientists should leverage data logs from smartphones to study human mobility, instead of relying on out-dated pen-and-paper questionnaires. Now we will explore the practical difficulties in using smartphone location logs. Finally we will introduce Personal Map Matched Imputation (PPMI), a method for surmounting these difficulties. We will compare PPMI to existing methods in the literature.

# Background

Smartphone location measures are obtained primarily (but not exclusively) by Global Positioning System (GPS) measurements. A GPS sensor uses the distance between a device and several satellites to determine the location of the device. Although using a GPS sensors is the most accurate way to establish location on a smartphone, the GPS sensor is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In order to avoid battery depletion and to overcome computational constraints smartphones also use less-accurate heuristics such as WiFi access points and cellphone triangulation.

However, not all GPS measurements are made equal. It is useful to classify GPS logs into different categories depending on their origin because different types of logs pose different methodological problems. We argue that social science researchers should explore using automatically collected smartphone GPS logs.  In most studies participants receive a specialised GPS devices to track their movement. We call resulting logs *specialised logs*.

@barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). The problem with this is that researchers have to develop a custom-made app, distribute it among research participants and enforce participation. Participants tend not to like these tracking apps because they drain the battery of the device.

Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) . For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017]. Often, secondary logs span several years. By law, secondary logs are accessible to users for free [@commission_protecting_2017]. Yet, secondary logs also present methodological challenges. They were created for non-academic purposes under engineering constraints ( these constraints will be detailed subsequently). These constraints mean that sensors do not track users continuously, meaning that the resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with missing data and measurement noise.

@@

Another methodological problem is related to the noise in the measurements that are collected. The accuracy of smartphone location measurements is substantially lower than that of professional GPS trackers. Android phones collect location information through a variety of methods.

*removed somethign here*

 In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time.

@@
